---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/README-",
  dpi = 250
)
```

# __m2r__

__m2r__ is a very new package that provides a persistent connection between [R](https://www.r-project.org) and [Macaulay2](http://www.math.uiuc.edu/Macaulay2/).

The package grew out of a collaboration at the algebraic statistics [2016 Mathematics Research Community](http://www.ams.org/programs/research-communities/mrc-16) and is currently being actively developed.



## Getting started

__m2r__ is loaded like any other R package:
```{r}
library(m2r)
```

When loaded, __m2r__ initializes a persistent connection to a back-end Macaulay2 session.  The basic function in R that accesses this connection is `m2()`, which simply accepts a character string that is run by the Macaulay2 session.

```{r}
m2("1 + 1")
```

You can see the persistence by setting variables and accessing them across different `m2()` calls:

```{r}
m2("x = 1")
m2("x")
```



## Rings, ideals, and Grobner bases




## Installation

Here's how you can install this _very developmental_ version of __m2r__.  Remember you need to have [Macaulay2](http://www.math.uiuc.edu/Macaulay2/) downloaded; __m2r__ will look for it in your path variable (in the terminal, `echo $PATH`) as set by `~/.bash_profile` or, if nonexistent, then `~/.bashrc`, then `~/.profile`.

```R
# install.packages("devtools")
devtools::install_github("musicman3320/m2r")
```
